Endogenous Macrodynamics in Algorithmic Recourse

### 2022-08-17 paper/paper.tex
more polishing

\IEEEoverridecommandlockouts
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithm}% http://ctan.org/pkg/algorithms
\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{placeins}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

@@

\caption{Dynamics in Algorithmic Recourse: we have a simple linear classifier trained for binary classification (a); the implementation of AR for a random subset of individuals leads to a noticable domain shift (b); as the classifier is retrained we observe a corresponding model shift (c); as this process is repeated, the decision boundary moves away from the target class (d).}\label{fig:poc}
\end{figure}

@@

\begin{equation}
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\text{yloss}(M(f(s_k^\prime)),y^*)}+ \lambda {\text{cost}(f(s_k^\prime)) }  \right\} \label{eq:general}
\end{aligned} 
\end{equation}

@@

Finally, DiCE \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} distinguishes itself from all other generators considered here in that it aims to generate a diverse set of \(K>1\) counterfactuals. Wachter et al. \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} show that diverse outcomes can in principal be achieved simply rerunning counterfactual search multiple times using stochastic gradient descent (or by randomly initializing the counterfactual). In \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function \(\text{cost}(\mathbf{s}^\prime)\) and therebt produce counterfactuals \(s_1, ... , s_K\) that look as different from each other as possible. The implementation of DiCE in the our library of choice - \texttt{CounterfactualExplanations.jl} - uses that exact approach. It is worth noting that for \(k=1\), DiCE reduces to Wachter since the DDP is constant and therefore does not affect the objective function in Equation \eqref{eq:general}.

@@

\begin{equation}
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\text{yloss}(M(f(s_k^\prime)),y^*)} \\ &+ \lambda_1 {\text{cost}(f(s_k^\prime))} + \lambda_2 {\text{extcost}(f(s_k^\prime))}  \label{eq:collective}
\end{aligned} 
\end{equation}

Here \(\text{cost}(f(s_k^\prime))\) denotes the proxy for private costs faced by the individual as before and \(\lambda_1\) governs to what extent that private cost ought to be penalized. The newly introduced term \(\text{extcost}(f(s_k^\prime))\) is meant to capture and address external costs incurred by the collective of individuals in response to changes in \(\mathbf{s}^\prime\). The underlying concept of private and external costs is borrowed from Economics and well-established in that field: when the decisions or actions by some individual market participant generate external costs, then the market is said to suffer from negative externalities and considered inefficient \protect\hyperlink{ref-pindyck2014microeconomics}{{[}24{]}}. We think that this concept describes the endogenous dynamics of algorithmic recourse oberserved here very well. As with individual recourse, the exact choice of \(\text{extcost}(\cdot)\) is not obvious, nor do we intend to provide a definite answer in this work, if such even exists. That being said, we do propose a few potential mitigation strategies in Section \ref{mitigate}.

@@

The dynamics illustrated in Figure \ref{fig:poc} in Section \ref{intro} were generated through a simple experiment that aims to simulate the process of Algorithmic Recourse in practice. We begin in the static setting at time time \(t=0\): given some classifier \(M\) that was pre-trained on data \(\mathcal{D}=\mathcal{D}_0 \cup \mathcal{D}_1\) we generate recourse for a random batch of \(B\) individuals in the non-target class (\(\mathcal{D}_0\)). Note that we focus our attention on classification problems, since classification poses the most common practical use-case for Algorithmic Recourse. In order to simulate the dynamic process, we suppose that the model \(M\) is retrained following the actual implementation of recourse in time \(t=0\). Following the update to the model, we assume that at time \(t=1\) recourse is generated for yet another random subset of individuals in the non-target class. This process is repeated for a number of time periods \(T\). To get a clean read on endogenous dynamics we keep the total population of samples closed: we allow existing samples to move from factual to counterfactual states, but do not allow any entirely new samples to enter the population. The experimental setup is summarized in Algorithm \ref{algo-experiment}

\begin{algorithm}
\caption{Simulation Experiment}\label{algo-experiment}
\begin{algorithmic}[1]
\Procedure{Experiment}{$M,\mathcal{D},G$}
\State $t\gets 0$
\While{$t<T$}
\State $\mathcal{D}_B \subset \mathcal{D}_0$ \Comment{Sample from $\mathcal{D}_0$.}
\State $\mathcal{D}_B\gets G(\mathcal{D}_B)$ \Comment{Generate counterfactuals.}
\State $M\gets M(\mathcal{D})$ \Comment{Retrain model.}
\EndWhile
\State \textbf{return} $M,\mathcal{D}$
\EndProcedure
\end{algorithmic}
\end{algorithm}

A noteworthy practical consideration is the choice of \(T\) and \(B\). The higher these values, the more factual instances undergo recourse throughout the entire experiment. Of course, this is likely to lead to more pronounced domain and model shifts by time \(T\). At the same time, it is generally improbable that a very large part of the population would request an explanation of the algorithm's decisions. In our experiments, we choose the values such that \(T \cdot B\) corresponds to the application of recourse on \(\approx50\%\) of the negative instances from the initial dataset. As we collect data at each time \(t\), we can also verify the impact of recourse when it is implemented for a smaller number of individuals.

@@

For each dataset and generator we look at three different types of classifiers all of them built and trained using \texttt{Flux.jl} \protect\hyperlink{ref-innes2018fashionable}{{[}29{]}}: firstly, a simple linear classifier - \textbf{Logistic Regression} - implemented as single linear layer with sigmoid activation; secondly, a multilayer perceptron (\textbf{MLP}); and finally, a \textbf{Deep Ensemble} composed of five MLPs following \protect\hyperlink{ref-lakshminarayanan2016simple}{{[}30{]}} that serves as our only probabilistic classifier. We have chosen to work with deep ensembles both for their simplicity and effectiveness at modelling predictive uncertainty. They are also the model of choice in \protect\hyperlink{ref-schut2021generating}{{[}6{]}}. The actual neural network architectures are kept simple (Table \ref{tab:mlp}), since we are only marginally concerned with achieving good initial classifier performance. For the real-world datasets we using mini-batch training and dropout regularization.

The Latent Space generator relies on a separate generative model. Following the authors of both REVISE and CLUE we use Variational Autoencoders (\textbf{VAE}) for this purpose. As with the classifiers, we deliberately choose to work with fairly simple architectures (Table \ref{tab:vae}). More expressive generative models generally also lead to more meaningful counterfactuals produced by Latent Space generators. But in our view this should simply be considered as a vulnerability of counterfactual generators that rely on surrogate models to learn what realistic representations of the underlying data.

@@

\caption{\label{tab:mlp}MLP architecture and training parameters.}
\centering
\begin{tabular}[t]{lrrll}
\toprule
  & Hidden Dim. & Hidden Layers & Batch & Dropout\\
\midrule
Synthetic & 32 & 1 & - & -\\
Real-World & 32 & 2 & 50 & 0.25\\
\bottomrule
\end{tabular}
\end{table}

\caption{\label{tab:vae}VAE architecture and training parameters.}
\centering
\begin{tabular}[t]{lrr}
\toprule
  & Hidden Dim. & Epochs\\
\midrule
Synthetic & 2 & 100\\
Real-World & 8 & 250\\
\bottomrule
\end{tabular}
\end{table}

@@

We use four synthetic binary classification datasets consisting of 1000 samples each. The datasets are presented in Figure \ref{fig:synthetic-data} (see also Appendix A for a formal description). Samples from the negative class are marked in blue while samples of the positive class are marked in orange.

@@

\caption{Synthetic classification datasets used in our experiments.}\label{fig:synthetic-data}
\end{figure}

@@

@@

\begin{equation}
\begin{aligned}
\text{extcost}(f(s_k^\prime)) = l(M(f(s_k^\prime)),y^\prime) \label{eq:clap}
\end{aligned}
\end{equation}

@@

\begin{equation}
\begin{aligned}
\text{extcost}(f(s_k^\prime)) = \text{dist}(f(s_k^\prime),\bar{x})  \label{eq:grav}
\end{aligned}
\end{equation}

@@

\caption{Illustrative example demonstrating the properties of the various mitigation strategies.}\label{fig:mitigation}
\end{figure}

>>>>

\IEEEoverridecommandlockouts
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithm}% http://ctan.org/pkg/algorithms
\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{placeins}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

\caption{Dynamics in Algorithmic Recourse: (a) we have a simple linear classifier trained for binary classification where samples from the negative class ($y=0$) are marked in blue and samples of the positive class ($y=1$) are marked in orange; (b) the implementation of AR for a random subset of individuals leads to a noticable domain shift; (c) as the classifier is retrained we observe a corresponding model shift; (d) as this process is repeated, the decision boundary moves away from the target class.}\label{fig:poc}
\end{figure}

@@

\begin{equation}
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{  {\text{yloss}(M(f(\mathbf{s}^\prime)),y^*)}+ \lambda {\text{cost}(f(\mathbf{s}^\prime)) }  \right\} \label{eq:general}
\end{aligned} 
\end{equation}

@@

Finally, DiCE \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} distinguishes itself from all other generators considered here in that it aims to generate a diverse set of \(K>1\) counterfactuals. Wachter et al. \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} show that diverse outcomes can in principal be achieved simply rerunning counterfactual search multiple times using stochastic gradient descent (or by randomly initializing the counterfactual)\footnote{Note, in fact, that \eqref{eq:general} naturally lends itself to that idea: setting \(K\) to some value greater than one and using the Wachter objective essentially boils down to computing multiple counterfactuals in parallel. Here, \(yloss(\cdot)\) is first broadcasted over elements of \(\mathbf{s}^\prime\) and then aggregated. This is exactly how counterfactual search is implemented in \texttt{CounterfactualExplanations.jl}.}. In \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function \(\text{cost}(\mathbf{s}^\prime)\) and thereby produce counterfactuals \(s_1, ... , s_K\) that look as different from each other as possible. The implementation of DiCE in the our library of choice - \texttt{CounterfactualExplanations.jl} - uses that exact approach. It is worth noting that for \(k=1\), DiCE reduces to Wachter since the DDP is constant and therefore does not affect the objective function in Equation \eqref{eq:general}.

@@

\begin{equation}
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \{ {\text{yloss}(M(f(\mathbf{s}^\prime)),y^*)} \\ &+ \lambda_1 {\text{cost}(f(\mathbf{s}^\prime))} + \lambda_2 {\text{extcost}(f(\mathbf{s}^\prime))} \}  \label{eq:collective}
\end{aligned} 
\end{equation}

Here \(\text{cost}(f(\mathbf{s}^\prime))\) denotes the proxy for private costs faced by the individual as before and \(\lambda_1\) governs to what extent that private cost ought to be penalized. The newly introduced term \(\text{extcost}(f(\mathbf{s}^\prime))\) is meant to capture and address external costs incurred by the collective of individuals in response to changes in \(\mathbf{s}^\prime\). The underlying concept of private and external costs is borrowed from Economics and well-established in that field: when the decisions or actions by some individual market participant generate external costs, then the market is said to suffer from negative externalities and considered inefficient \protect\hyperlink{ref-pindyck2014microeconomics}{{[}24{]}}. We think that this concept describes the endogenous dynamics of algorithmic recourse observed here very well. As with individual recourse, the exact choice of \(\text{extcost}(\cdot)\) is not obvious, nor do we intend to provide a definite answer in this work, if such even exists. That being said, we do propose a few potential mitigation strategies in Section \ref{mitigate}.

@@

The dynamics illustrated in Figure \ref{fig:poc} in Section \ref{intro} were generated through a simple experiment that aims to simulate the process of Algorithmic Recourse in practice. We begin in the static setting at time time \(t=0\): firstly, we have some binary classifier \(M\) that was pre-trained on data \(\mathcal{D}=\mathcal{D}_0 \cup \mathcal{D}_1\), where \(\mathcal{D}_0\) and \(\mathcal{D}_1\) denote samples in the non-target and target class, respectively; secondly, we generate recourse for a random batch of \(B\) individuals in the non-target class (\(\mathcal{D}_0\)). Note that we focus our attention on classification problems, since classification poses the most common practical use-case for Algorithmic Recourse\footnote{To keep notation simple we have also restricted ourselves to binary classification here, but \texttt{AlgorithmicRecourseDynamics.jl} can also be used for multi-class problems.}.

In order to simulate the dynamic process, we suppose that the model \(M\) is retrained following the actual implementation of recourse in time \(t=0\). Following the update to the model, we assume that at time \(t=1\) recourse is generated for yet another random subset of individuals in the non-target class. This process is repeated for a number of time periods \(T\). To get a clean read on endogenous dynamics we keep the total population of samples closed: we allow existing samples to move from factual to counterfactual states, but do not allow any entirely new samples to enter the population. The experimental setup is summarized in Algorithm \ref{algo-experiment}

\begin{algorithm}
\caption{Simulation Experiment}\label{algo-experiment}
\begin{algorithmic}[1]
\Procedure{Experiment}{$M,\mathcal{D},G$}
\State $E\gets \emptyset$ \Comment{Initialize evaluation $E$.}
\State $t\gets 0$
\While{$t<T$}
\State $\text{batch} \subset \mathcal{D}_0$ \Comment{Sample from $\mathcal{D}_0$  (assignment).}
\State $\text{batch}\gets G(\text{batch})$ \Comment{Generate counterfactuals.}
\State $M\gets M(\mathcal{D})$ \Comment{Retrain model.}
\State $E\gets \text{eval}(M,\mathcal{D}) \cup E$ \Comment{Update evaluation.}
\State $t\gets t+1$ \Comment{Increment $t$.}
\EndWhile
\State \textbf{return} $E, M,\mathcal{D}$
\EndProcedure
\end{algorithmic}
\end{algorithm}

Note that the operation in line 4 is an assignment, rather than a copy operation, so any updates to `batch' will also affect \(\mathcal{D}\). The function \(\text{eval}(M,\mathcal{D})\) loosely denotes the computation of various evaluation metrics introduced below. In practice, these metrics can also be computed at regular intervals as opposed to every round.

Along with any other fixed parameters affecting the counterfactual search, the parameters \(T\) and \(B\) are assumed as given in Algorithm \ref{algo-experiment}. Still, it worth noting that the higher these values, the more factual instances undergo recourse throughout the entire experiment. Of course, this is likely to lead to more pronounced domain and model shifts by time \(T\). In our experiments, we choose the values such that \(T \cdot B\) corresponds to the application of recourse on \(\approx50\%\) of the negative instances from the initial dataset. As we compute evaluation metrics at regular intervals throughout the procedure, we can also verify the impact of recourse when it is implemented for a smaller number of individuals.

@@

For each dataset and generator we look at three different types of classifiers all of them built and trained using \texttt{Flux.jl} \protect\hyperlink{ref-innes2018fashionable}{{[}29{]}}: firstly, a simple linear classifier - \textbf{Logistic Regression} - implemented as single linear layer with sigmoid activation; secondly, a multilayer perceptron (\textbf{MLP}); and finally, a \textbf{Deep Ensemble} composed of five MLPs following \protect\hyperlink{ref-lakshminarayanan2016simple}{{[}30{]}} that serves as our only probabilistic classifier. We have chosen to work with deep ensembles both for their simplicity and effectiveness at modelling predictive uncertainty. They are also the model of choice in \protect\hyperlink{ref-schut2021generating}{{[}6{]}}. The actual neural network architectures are kept simple (top half of Table \ref{tab:architecture}), since we are only marginally concerned with achieving good initial classifier performance.

The Latent Space generator relies on a separate generative model. Following the authors of both REVISE and CLUE we use Variational Autoencoders (\textbf{VAE}) for this purpose. As with the classifiers, we deliberately choose to work with fairly simple architectures (bottom half of Table \ref{tab:architecture}). More expressive generative models generally also lead to more meaningful counterfactuals produced by Latent Space generators. But in our view this should simply be considered as a vulnerability of counterfactual generators that rely on surrogate models to learn what realistic representations of the underlying data.

@@

\caption{\label{tab:architecture}Neural network architectures and training parameters.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{llrlrllr}
\toprule
 & Data & Hidden Dim. & Latent Dim. & Hidden Layers & Batch & Dropout & Epochs\\
\midrule
\addlinespace[0.3em]
\multicolumn{8}{l}{\textbf{MLP}}\\
\hspace{1em} & Synthetic & 32 & - & 1 & - & - & 100\\
\cmidrule{2-8}
\hspace{1em} & Real-World & 32 & - & 2 & 50 & 0.25 & 100\\
\cmidrule{1-8}
\addlinespace[0.3em]
\multicolumn{8}{l}{\textbf{VAE}}\\
\hspace{1em} & Synthetic & 32 & 2 & 1 & - & - & 100\\
\cmidrule{2-8}
\hspace{1em} & Real-World & 32 & 8 & 1 & - & - & 250\\
\bottomrule
\end{tabular}}
\end{table}

@@

We use four synthetic binary classification datasets consisting of 1000 samples each: \textbf{Overlapping}, \textbf{Linearly Separable}, \textbf{Circles} and \textbf{Moons}. The datasets are presented in Figure \ref{fig:synthetic-data}. Samples from the negative class are marked in blue while samples of the positive class are marked in orange.

@@

\caption{Synthetic classification datasets used in our experiments. Samples from the negative class ($y=0$) are marked in blue while samples of the positive class ($y=1$) are marked in orange.}\label{fig:synthetic-data}
\end{figure}

@@

To recap, we can answer our research questions as follows: firstly, endogenous dynamics do emerge in our experiments and we find them substantial enough to be considered costly; secondly, the choice of the counterfactual generator does matter, with Latent Space search generally having a dampening effect. The observed dynamics therefore seem to be driven by a discrepancy between counterfactual outcomes that minimize costs to the individual and outcomes that comply with the data generating process.

@@

\begin{equation}
\begin{aligned}
\text{extcost}(f(\mathbf{s}^\prime)) = l(M(f(\mathbf{s}^\prime)),y^\prime) \label{eq:clap}
\end{aligned}
\end{equation}

@@

\begin{equation}
\begin{aligned}
\text{extcost}(f(\mathbf{s}^\prime)) = \text{dist}(f(\mathbf{s}^\prime),\bar{x})  \label{eq:grav}
\end{aligned}
\end{equation}

@@

\caption{Illustrative example demonstrating the properties of the various mitigation strategies. Samples from the negative class ($y=0$) are marked in blue while samples of the positive class ($y=1$) are marked in orange.}\label{fig:mitigation}
\end{figure}

### 2022-08-17 paper/sections/empirical.rmd
more polishing

For each dataset and generator we look at three different types of classifiers all of them built and trained using `Flux.jl` [@innes2018fashionable]: firstly, a simple linear classifier - **Logistic Regression** - implemented as single linear layer with sigmoid activation; secondly, a multilayer perceptron (**MLP**); and finally, a **Deep Ensemble** composed of five MLPs following @lakshminarayanan2016simple that serves as our only probabilistic classifier. We have chosen to work with deep ensembles both for their simplicity and effectiveness at modelling predictive uncertainty. They are also the model of choice in @schut2021generating. The actual neural network architectures are kept simple (Table \@ref(tab:mlp)), since we are only marginally concerned with achieving good initial classifier performance. For the real-world datasets we using mini-batch training and dropout regularization.

The Latent Space generator relies on a separate generative model. Following the authors of both REVISE and CLUE we use Variational Autoencoders (**VAE**) for this purpose. As with the classifiers, we deliberately choose to work with fairly simple architectures (Table \@ref(tab:vae)). More expressive generative models generally also lead to more meaningful counterfactuals produced by Latent Space generators. But in our view this should simply be considered as a vulnerability of counterfactual generators that rely on surrogate models to learn what realistic representations of the underlying data.

```{r mlp}
tab <- data.frame(
  "Hidden Dim." = c(32,32),
  "Hidden Layers." = c(1,2),
  "Batch" = c("-",50),
  "Dropout" = c("-",0.25)
)
rownames(tab) <- c("Synthetic", "Real-World")
knitr::kable(
  tab, booktabs = TRUE,
  caption = 'MLP architecture and training parameters.',
  col.names = c("Hidden Dim.","Hidden Layers", "Batch", "Dropout")
)
```

```{r vae}
tab <- data.frame(
  "Hidden Dim." = c(2,8),
  "Epochs" = c(100,250)
)
rownames(tab) <- c("Synthetic", "Real-World")
knitr::kable(
  tab, booktabs = TRUE,
  caption = 'VAE architecture and training parameters.',
  col.names = c("Hidden Dim.","Epochs")
)
```

@@

We use four synthetic binary classification datasets consisting of 1000 samples each. The datasets are presented in Figure \@ref(fig:synthetic-data) (see also Appendix A for a formal description). Samples from the negative class are marked in blue while samples of the positive class are marked in orange.

```{r synthetic-data, fig.cap="Synthetic classification datasets used in our experiments."}
knitr::include_graphics("www/synthetic_data.png")
```

>>>>

For each dataset and generator we look at three different types of classifiers all of them built and trained using `Flux.jl` [@innes2018fashionable]: firstly, a simple linear classifier - **Logistic Regression** - implemented as single linear layer with sigmoid activation; secondly, a multilayer perceptron (**MLP**); and finally, a **Deep Ensemble** composed of five MLPs following @lakshminarayanan2016simple that serves as our only probabilistic classifier. We have chosen to work with deep ensembles both for their simplicity and effectiveness at modelling predictive uncertainty. They are also the model of choice in @schut2021generating. The actual neural network architectures are kept simple (top half of Table \@ref(tab:architecture)), since we are only marginally concerned with achieving good initial classifier performance.

The Latent Space generator relies on a separate generative model. Following the authors of both REVISE and CLUE we use Variational Autoencoders (**VAE**) for this purpose. As with the classifiers, we deliberately choose to work with fairly simple architectures (bottom half of Table \@ref(tab:architecture)). More expressive generative models generally also lead to more meaningful counterfactuals produced by Latent Space generators. But in our view this should simply be considered as a vulnerability of counterfactual generators that rely on surrogate models to learn what realistic representations of the underlying data.

```{r architecture}
tab <- data.frame(
  "Model" = c("MLP","MLP","VAE","VAE"),
  "Data" = c("Synthetic", "Real-World", "Synthetic", "Real-World"),
  "Hidden" = c(32,32,32,32),
  "Latent" = c("-","-",2,8),
  "Layers" = c(1,2,1,1),
  "Batch" = c("-",50,"-","-"),
  "Dropout" = c("-",0.25,"-","-"),
  "Epochs" = c(100,100,100,250)
)
library(kableExtra)
kbl(
  tab, booktabs = TRUE,
  caption = 'Neural network architectures and training parameters.',
  col.names = c("Model","Data","Hidden Dim.","Latent Dim.","Hidden Layers", "Batch", "Dropout", "Epochs")
) |> 
  collapse_rows(1:2, row_group_label_position = 'stack') |>
  kable_styling(latex_options = c("scale_down"))
```

@@

We use four synthetic binary classification datasets consisting of 1000 samples each: **Overlapping**, **Linearly Separable**, **Circles** and **Moons**. The datasets are presented in Figure \@ref(fig:synthetic-data). Samples from the negative class are marked in blue while samples of the positive class are marked in orange.

```{r synthetic-data, fig.cap="Synthetic classification datasets used in our experiments. Samples from the negative class ($y=0$) are marked in blue while samples of the positive class ($y=1$) are marked in orange."}
knitr::include_graphics("www/synthetic_data.png")
```

### 2022-08-17 paper/sections/methodology.rmd
more polishing

\begin{equation}
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\text{yloss}(M(f(s_k^\prime)),y^*)}+ \lambda {\text{cost}(f(s_k^\prime)) }  \right\} (\#eq:general)
\end{aligned} 
\end{equation}

@@

Finally, DiCE [@mothilal2020explaining] distinguishes itself from all other generators considered here in that it aims to generate a diverse set of $K>1$ counterfactuals. Wachter et al. [@wachter2017counterfactual] show that diverse outcomes can in principal be achieved simply rerunning counterfactual search multiple times using stochastic gradient descent (or by randomly initializing the counterfactual). In @mothilal2020explaining diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function $\text{cost}(\mathbf{s}^\prime)$ and therebt produce counterfactuals $s_1, ... , s_K$ that look as different from each other as possible. The implementation of DiCE in the our library of choice - `CounterfactualExplanations.jl` - uses that exact approach. It is worth noting that for $k=1$, DiCE reduces to Wachter since the DDP is constant and therefore does not affect the objective function in Equation \@ref(eq:general).

@@

\begin{equation}
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\text{yloss}(M(f(s_k^\prime)),y^*)} \\ &+ \lambda_1 {\text{cost}(f(s_k^\prime))} + \lambda_2 {\text{extcost}(f(s_k^\prime))}  (\#eq:collective)
\end{aligned} 
\end{equation}

Here $\text{cost}(f(s_k^\prime))$ denotes the proxy for private costs faced by the individual as before and $\lambda_1$ governs to what extent that private cost ought to be penalized. The newly introduced term $\text{extcost}(f(s_k^\prime))$ is meant to capture and address external costs incurred by the collective of individuals in response to changes in $\mathbf{s}^\prime$. The underlying concept of private and external costs is borrowed from Economics and well-established in that field: when the decisions or actions by some individual market participant generate external costs, then the market is said to suffer from negative externalities and considered inefficient [@pindyck2014microeconomics]. We think that this concept describes the endogenous dynamics of algorithmic recourse oberserved here very well. As with individual recourse, the exact choice of $\text{extcost}(\cdot)$ is not obvious, nor do we intend to provide a definite answer in this work, if such even exists. That being said, we do propose a few potential mitigation strategies in Section \@ref(mitigate).

>>>>

\begin{equation}
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{  {\text{yloss}(M(f(\mathbf{s}^\prime)),y^*)}+ \lambda {\text{cost}(f(\mathbf{s}^\prime)) }  \right\} (\#eq:general)
\end{aligned} 
\end{equation}

@@

Finally, DiCE [@mothilal2020explaining] distinguishes itself from all other generators considered here in that it aims to generate a diverse set of $K>1$ counterfactuals. Wachter et al. [@wachter2017counterfactual] show that diverse outcomes can in principal be achieved simply rerunning counterfactual search multiple times using stochastic gradient descent (or by randomly initializing the counterfactual)^[Note, in fact, that \@ref(eq:general) naturally lends itself to that idea: setting $K$ to some value greater than one and using the Wachter objective essentially boils down to computing multiple counterfactuals in parallel. Here, $yloss(\cdot)$ is first broadcasted over elements of $\mathbf{s}^\prime$ and then aggregated. This is exactly how counterfactual search is implemented in `CounterfactualExplanations.jl`.]. In @mothilal2020explaining diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function $\text{cost}(\mathbf{s}^\prime)$ and thereby produce counterfactuals $s_1, ... , s_K$ that look as different from each other as possible. The implementation of DiCE in the our library of choice - `CounterfactualExplanations.jl` - uses that exact approach. It is worth noting that for $k=1$, DiCE reduces to Wachter since the DDP is constant and therefore does not affect the objective function in Equation \@ref(eq:general).

@@

\begin{equation}
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \{ {\text{yloss}(M(f(\mathbf{s}^\prime)),y^*)} \\ &+ \lambda_1 {\text{cost}(f(\mathbf{s}^\prime))} + \lambda_2 {\text{extcost}(f(\mathbf{s}^\prime))} \}  (\#eq:collective)
\end{aligned} 
\end{equation}

Here $\text{cost}(f(\mathbf{s}^\prime))$ denotes the proxy for private costs faced by the individual as before and $\lambda_1$ governs to what extent that private cost ought to be penalized. The newly introduced term $\text{extcost}(f(\mathbf{s}^\prime))$ is meant to capture and address external costs incurred by the collective of individuals in response to changes in $\mathbf{s}^\prime$. The underlying concept of private and external costs is borrowed from Economics and well-established in that field: when the decisions or actions by some individual market participant generate external costs, then the market is said to suffer from negative externalities and considered inefficient [@pindyck2014microeconomics]. We think that this concept describes the endogenous dynamics of algorithmic recourse observed here very well. As with individual recourse, the exact choice of $\text{extcost}(\cdot)$ is not obvious, nor do we intend to provide a definite answer in this work, if such even exists. That being said, we do propose a few potential mitigation strategies in Section \@ref(mitigate).
